Mechanisation of Session-based Lambda Calculus Type Systems

### 2014-12-04 proposal/proposal.tex
Added (some) work plan, session types disc. and fixed "binds" anomaly.

@@

@@

@@

In the protocol defined above, a client selects between buying a book or
receiving a price quote for a book. In the buy option case, the client sends
the name of the book to the server and their card details, then the server
sends the receipt for the purchase. In the quote option case, the client sends
the name of the book and receives a price quote from the server. I have
replaced the types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more
meaningful type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate;
all are synonyms for the type $\1$ which denotes the terminal session
type. The client and server process bodies are as follows:
\begin{gather*}
\sdef{ClientBody_c}
     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
\sdef{ServerBody_c}
     {\CPcase{c}
             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
\end{gather*}

@@

The conclusion binds a variable in (C) but a type in (CCO). Thus, in the
premise for (CCO) one can choose a suitably fresh $x$ in each premise. Note
that, $L$ can be chosen such that $\Psi \cup \Phi \cup \fv(M) \cup \fv(N)
\subseteq L$, where $\fv(M)$ are the set of free variables in term $M$, and
$x$ can be the same in both premises. Similar changes are made to the rules
for CP so that, for example, output which allocates a fresh variable $y$ no
longer names $y$ in the conclusion: \input{cp-output}

@@

I aim to have completed the GV and CP definitions including reduction and
commuting conversion rules for CP by the end of December. This will require
definition of infrastructure for handling binders for both GV, CP channels and
CP propositional variables, and inductive definitions for well-typed terms,
commuting conversions and CP principal cut reductions. So far I have completed
some of the definitions for well-typed terms and CP process rules; the
remaining rules require the infrastructure for substitution and opening of
binders.

After December, it remains to define the translation between GV and CP which I
envisage will be the bulk of the work, requiring several rules on typing
derivations; I plan to define these as inductive datatype constructors but I
shall take care to reuse any suitable libraries that are available, e.g. a
library for handling CPS translations. For this step of the work, I allocated
one month but this is a cautious estimate accounting some slack time for any
delays in preceding stages of the development; I suspect this stage will not
require a full month.

For the remainder, I shall look to extend the definitions of GV and CP using
the definitions I have provided. As discussed previously, it would be
interesting to study a system for supporting side-effecting functions (aka
closures) for session channels. In order to achieve this, operational
semantics for GV would need to be defined, and the recent draft by
\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
do this~\cite{Lindley:2014:SPS}. I allocated the remaining time (approx. two
months) to perform this extension because I believe it requires the most
development: extending existing definitions to handle side-effecting
functions, defining operational semantics, proving subject reduction and
related properties. Additionally, it will also require a theoretical treatment
prior to undertaking the development task, in contrast to the preceding
development work which mostly utilises results from the literature.

>>>>

\usepackage{tabularx}
\usepackage{tabulary}

@@

\newcommand{\Deliverable}[2]
           {\multicolumn{2}{|c|}{\textbf{D{#1}: {#2}}} \\ \hline}
\newcommand{\DSubD}[3]{D{#1}.{#2} & #3 \\ \hline}
\newcommand{\DTask}[6]{D#1T#2 & #3 & #4 & #5 & #6 \\ \hline}
\newcommand{\IDTsk}[2]{Task D{#1}T{#2}}

@@

\subsubsection{Preamble}

The definition of session types will be important for the remainder:
\input{session-types.tex}

Here, I use a modified version of the definition given by
\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} where I have omitted recursive
session types. The type \lstinline{end} denotes the terminal session type and
a channel typed as such may not perform any communication. A communication
channel with session type $\inpt{T}{S}$ expects to receive a term of type $T$
and then continues with session type $S$. Conversely, a channel with type
$\outpt{T}{S}$ can send a term of type $T$ and then has type $S$. Thus, input
and output are considered \emph{dual} session types, in the sense that a
channel with one endpoint having type $\inpt{T}{S}$ and the other endpoint
having type $\outpt{T}{\Sbar{S}}$, where $\Sbar{S}$ is the dual of $S$,
permits communication between two agents, allowing one to send a term of type
$T$ and the other to receive it with continuing endpoint types $\Sbar{S}$ and
$S$, respectively. Similarly, the remaining types $\Branch{l_i}{S_i}{i\in I}$
and $\Choice{l_i}{S_i}{i\in I}$ are duals of each other. A channel of type
$\Branch{l_i}{S_i}{i\in I}$ can receive a label $l_i$ which will select the
corresponding session type $S_i$ as the continuing type of the channel. While
a channel with type $\Choice{l_i}{S_i}{i\in I}$ can send a label $l_i$ and
continues as $S_i$.

@@

Note, I am using $\tensor{}{}$ for session output and $\lfn{}{}$ for session
input following the presentation by \citeauthor{Caires:2010:STI}. In the
protocol defined above, a client selects between buying a book or receiving a
price quote for a book. In the buy option case, the client sends the name of
the book to the server and their card details, then the server sends the
receipt for the purchase. In the quote option case, the client sends the name
of the book and receives a price quote from the server. I have replaced the
types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more meaningful
type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate; all are
synonyms for the type $\1$ which denotes the terminal session type. The client
and server process bodies are as follows:
\begin{gather*}
\sdef{ClientBody_c}
     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
\sdef{ServerBody_c}
     {\CPcase{c}
             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
\end{gather*}

@@

The conclusion specifies a variable in (C) but a type in (CCO). Thus, in the
premise for (CCO) one can choose a suitably fresh $x$ in each premise. Note
that, $L$ can be chosen such that $\Psi \cup \Phi \cup \fv(M) \cup \fv(N)
\subseteq L$, where $\fv(M)$ are the set of free variables in term $M$, and
$x$ can be the same in both premises. Similar changes are made to the rules
for CP so that, for example, output which allocates a fresh variable $y$ no
longer names $y$ in the conclusion: \input{cp-output}

@@

The deliverables can be split into three main areas as indicated in the
following table:

\begin{tabularx}{\linewidth}{|c|X|}\hline
Deliverable \# &  Description \\ \hline
\Deliverable{1}{Mechanisation of Wadler's system}
\DSubD{1}{*}{Mechanisation of Wadler's system}
\Deliverable{2}{Closure side-effecting extension}
\DSubD{2}{1}{Implementation of the GV operational semantics}
\DSubD{2}{2}{Implementation of the closure side-effecting extension}
\Deliverable{3}{Report write-up}
\DSubD{3}{1}{Description of approach, Wadler mechanisation and related work}
\end{tabularx}

The current proposal will form part of D3.1 (related work, etc.). The
remaining parts of D3 will be written up as the project progresses, conducted
in tandem with D1 and D2. The estimated work plan is given in the following
table and sub-divides the above deliverable stages into tasks. The tasks
marking the completion of a deliverable are noted. The tasks appear in
chronological order.

\begin{center}
\begin{tabular}{|*{5}{c|}}\hline
\multicolumn{5}{|c|}{\textbf{Estimated Work Plan}}\\ \hline
Task & Start Date & Duration (days) & End Date & Deliverables \\ \hline
\DTask{1}{1}{8th December}{15}{23rd December}{-}
\DTask{3}{1}{20th December}{3}{23rd December}{-}
\DTask{1}{2}{24th December}{17}{10th January}{D1}
\DTask{3}{2}{7th January}{11}{18th January}{D3.1}
\DTask{2}{1}{11th January}{7}{18th January}{-}
\DTask{2}{2}{19th January}{18}{6th Feburary}{D2.1}
%\DTask{3}{3}{
\end{tabular}
\end{center}

\IDTsk{1}{1} involves completing the GV and CP definitions including reduction
and commuting conversion rules for CP which I aim to complete during
December. This will require definition of infrastructure for handling binders
for both GV, CP channels and CP propositional variables, and inductive
definitions for well-typed terms, commuting conversions and CP principal cut
reductions. So far I have completed some of the definitions for well-typed
terms and CP process rules; the remaining rules require the infrastructure for
substitution and opening of binders. \IDTsk{3}{1} will consist of an initial
structuring and write up of the mechanisation effort of Wadler's system. It
will include a detailed description of the approach taken (which can, in part,
be formed from the current proposal) and the related work (again, partly
formed from the current document). Additionally, it will describe any issues
which have arisen, including changes made to any of the definitions as
presented by Wadler, including the changes already detailed here.

For D1, it remains to define the translation between GV and CP which I
envisage will require several rules on typing derivations; I plan to define
these as inductive datatype constructors but I shall take care to reuse any
suitable libraries that are available, e.g. a library for handling CPS
translations. For this step of the work (\IDTsk{1}{2}), I made a cautious
estimate on duration, accounting some slack time for any delays in preceding
stages of the development. \IDTsk{3}{2} will document the translation phase
and will also be used to document the initial approach for the extension (see
table for overlapping task allocation). \IDTsk{3}{2} represents delivery of
the first draft (D3.1) of the report describing the mechanisation of Wadler's
system but omitting the theoretical treatment for the proposed extension.

For the remainder, I shall look to extend the definitions of GV and CP using
the definitions I have provided. As discussed previously, it would be
interesting to study a system for supporting side-effecting functions (aka
closures) for session channels. In order to achieve this, operational
semantics for GV would need to be defined and the recent draft by
\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
do this~\cite{Lindley:2014:SPS}. The task allocation time indicated in the
work plan table. \IDTsk{2}{1} is a theoretical investigation into the
operational semantics, deciding whether the system by
\citeauthor{Lindley:2014:SPS} offers a suitable basis for the extension, and
investigating the details of the side-effecting closure extension (assuming an
appropriate base operational semantics for GV). \IDTsk{2}{2} is then the
implementation of the suitable operational semantics (without side-effecting
closure), culminating in the first part of the second deliverable
(D2.1). Lastly, \IDTsk{2}{3} is an allocation of time to the task of
implementing the extension permitting closures with side-effects. In my work
plan, I have assumed \IDTsk{2}{2} implemented an operational semantics that
eases this particular extension which justifies the task allocation time. Both
\IDTsk{2}{2} and \IDTsk{2}{3} may also include revisions to the theoretical
treatment started in \IDTsk{2}{1} as a consequence of initial design
oversights.
